Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2022-07-25 replication_paper/pnas_draft/paper.qmd

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
pnas_type: pnasresearcharticle

bibliography: ICBintro.bib
csl: pnas.csl

@@

#https://bookdown.org/yihui/bookdown/a-single-document.html
#https://stackoverflow.com/questions/52531637/knitr-rmarkdown-latex-how-to-cross-reference-figures-and-tables-in-2-different/52532269#52532269
#https://stackoverflow.com/questions/51595939/bookdown-cross-reference-figure-in-another-file
#https://stackoverflow.com/questions/25824795/how-to-combine-two-rmarkdown-rmd-files-into-a-single-output/51521542#51521542
#https://stackoverflow.com/questions/64700865/using-r-rstudio-knit-to-pdf-how-to-pass-custom-flags-to-the-pdflatex-command

bibliography: arxiv.bib
biblio-style: unsrt
output: rticles::arxiv_article

#Disabled several the caused errors
header-includes:
#  - \usepackage{xr} \externaldocument{appendix} #External references from the appendix
#  - \usepackage{tinytex}
  - \usepackage[utf8]{inputenc}
  - \usepackage{pifont}
  - \usepackage{newunicodechar}
  - \newunicodechar{✓}{\ding{51}}
  - \newunicodechar{✗}{\ding{55}}
  - \usepackage{array}
  - \usepackage{ctable} # added for demo
#  - \usepackage{natbib} #added for latex citation within huxtable
#  - \usepackage{biblatex} #added for latex citation within huxtable
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{titlesec}
  - \usepackage[parfill]{parskip}
  - \usepackage{makecell}
  - \usepackage{graphicx}
#  - \usepackage{caption}
#  - \usepackage[capposition=top]{floatrow}
#  - \titleformat{\subsubsection}{\normalfont\normalsize\itshape}{\thesubsubsection}{1em}{}
#  - \titlespacing*{\subsubsection}{0pt}{3.25ex plus 1ex minus .2ex}{0ex plus .2ex}
#  - \DeclareUnicodeCharacter{00A0}{ }
  - \usepackage{setspace}
  - \usepackage{cellspace}
  - \setlength\cellspacetoplimit{0.8ex}
  - \renewcommand{\arraystretch}{0.8}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
#  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
#  - \usepackage{subfigure}
  - \usepackage{subcaption}
  - \usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
  - \usepackage{float}
#  - \usepackage{eso-pic,graphicx,transparent}
#  - \usepackage[perpage]{footmisc}
  - \usepackage[algo2e]{algorithm2e}
#  - \usepackage[backend = biber]{biblatex} #added in hopes of getting a bbl
#https://tex.stackexchange.com/questions/131646/algorithm2e-command-algorithm-already-defined
---

```{r, echo=F, cache=F, messages=F, results='hide', warnings=F, include=FALSE}
print(wd())
#International Crisis Behavior Events (ICBe) Dataset
knitr::opts_chunk$set(echo=F, cache=TRUE, messages=F, results='hide', warnings=F, include=FALSE) #fig.width = 8, fig.height = 8, message=F, warning=F, , results=F

#Execute these lines to get bbl files for the arxiv submission
options(tinytex.clean=F)

```

```{r, eval=T, echo=F, cache=F}

#library(pacman)
library(googlesheets4)
library(tidyverse)
library(janitor)
library(flextable)
set_flextable_defaults(fonts_ignore=TRUE) #Warning: fonts used in `flextable` are ignored because the `pdflatex` engine is used and not `xelatex` or `lualatex`. You can avoid this warning by using the `set_flextable_defaults(fonts_ignore=TRUE)` command or use a compatible engine by defining `latex_engine: xelatex` in the YAML header of the R Markdown document.
library(ftExtra)
options(tidyverse.quiet = TRUE)
options(gargle_oauth_email = TRUE)

#install.packages('stargazer')
#install.packages('kableExtra')
#! sh: 1: pdflatex: not found
#sudo apt-get install texlive-latex-recommended texlive-fonts-recommended

#! LaTeX Error: File `letltxmacro.sty' not found.
#sudo apt-get install texlive-latex-extra

#install.packages('tinytex')
#tinytex::install_tinytex()

```

```{r, echo=F}
#Notes
#https://tex.stackexchange.com/questions/83440/inputenc-error-unicode-char-u8-not-set-up-for-use-with-latex
#- \DeclareUnicodeCharacter{00A0}{ }

```

```{r, include=FALSE}
# options(tinytex.verbose = TRUE)
```

```{r, echo=FALSE}

`%>%` <- magrittr::`%>%`

```

```{r, echo=FALSE}

icb_long_clean <- readRDS(file=paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_clean.Rds"))
codings_long_agreement <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_agreement.Rds") )

ICBe_events_agreed <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_events_agreed.Rds")) 
dim(ICBe_events_agreed)
icb_events_long <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_events_agreed_long.Rds")) 
dim(icb_events_long)

#oh is it because break is reserved in R? lol
#icb_crises <- read_csv(file=paste0(here::here(), "/data_other_datasets/icb/icb1v14.csv")) %>% janitor::clean_names()
icb_crises  <- read.csv(paste0(here::here(), "/replication_data/in/icb1v14.csv")) %>% janitor::clean_names()

#somehow break of all things is a b roken name
#icb_actors <- read_csv(file=paste0(here::here(), "/data_other_datasets/icb/icb2v14.csv"))
icb_actors <- read.csv(paste0(here::here(), "/replication_data/in/icb2v14.csv"))

```

<!-- note: please start your introduction without including the word "Introduction" as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs. -->

If we could record every important interaction between countries in all of diplomacy, military conflict, and international political economy, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could in principle be unbounded, but the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured, and accordingly, massive effort is expended in social science attempting to record these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] With improvements in natural language processing, more open-ended efforts have sought to capture entire unstructured streams of international events drawn from news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017. On event-extraction from images and social-media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations. We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherCrisisEscalationWar2000; @wilkenfeldInterstateCrisesViolence2000; @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state-of-the-art event data collection efforts. We find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs is unrecognizable through the lens of current quantification efforts.

This is a measurement paper that makes the following argument --- there is a real-world unobserved latent concept known as international relations, we propose a method for systematically measuring it, we successfully apply this method producing a new large scale set of measurements, those measurements exhibit several desirable kinds of internal and external validity, and those measurements out-perform other existing approaches. The article organizes that argument into eight sections: task definition; corpus; priors/existing state of the art; ICBe coding process; internal consistency; case study selection; recall; and precision. A final section concludes.

Task Definition {.unnumbered}
========================================

We consider the measurement task of abstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine readable abstractive events. Formally, a historical episode, $H$, is demarcated by a period of time $[T_{start}, T_{end}] \in T$, a set of Players $p \in P$, and a set of behaviors they undertook during that time $b \in B$. International Relations, $IR$, is the system of regularities that govern the strategic interactions that world actors make during a historical episode, given their available options, preferences, beliefs, and expectations of choices made by others. We observe neither $H$ nor $IR$ directly. Rather the Historical Record, $HR$, produces documents $d \in D$ containing some relevant and true (as well as irrelevant and untrue) information about behaviors that were undertaken recorded in the form of unstructured natural language text. The task is to combine informative priors about $IR$ with an unstructured corpus $D$ to produce a series of structured discrete events, $e \in E$, that have high coverage, precision, and recall over what actually took place in history, $H$.

```{r cuban case study precision, eval=T, results='hide', echo=F, include=T, message=F, cache=F}

ICBe_events_agreed_markdown <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_events_agreed_markdown.Rds"))

temp <- ICBe_events_agreed_markdown %>%
        filter(crisno==196) %>%
        group_by(crisno, sentence_number_int_aligned) %>%
          filter(!duplicated(value_markdown %>% trimws() )) %>% #note there are dupe markdowns for some reason
        ungroup() %>%
        mutate(banding=sentence_number_int_aligned %% 2)
temp$is_subsstring <- NA
for(i in 1:nrow(temp)){
  temp$is_subsstring[i] <-   stringr::str_detect(temp$value_markdown %>% tolower() %>% str_replace_all('[^a-z0-9]',"") ,
                                                                                pattern = fixed(temp$value_markdown[i] %>% tolower() %>% str_replace_all('[^a-z0-9]',"") )
                                                                              ) %>% replace_na(0) %>%  sum()
}
temp <-temp %>%
        mutate(value_markdown = ifelse(is_subsstring<=1,value_markdown,'')) %>%
        group_by(sentence_number_int_aligned) %>%
          filter(value_markdown!='' | length(value_markdown)==1) %>%
        ungroup() %>%
        mutate(crisno= ifelse(!is.na(lag(sentence_span_text)) & sentence_span_text==lag(sentence_span_text), "", crisno)) %>%
        mutate(sentence_number_int_aligned= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_number_int_aligned)) %>%
        mutate(sentence_span_text= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_span_text))   #C=crisno, 
      #head(100) %>%
ft_cuba_precision <- temp %>%
                     dplyr::select(S=sentence_number_int_aligned, ICB=sentence_span_text, ICBe=value_markdown)  %>%
                      flextable::as_flextable() %>%
        flextable::set_header_labels(values =
        list(
              S = "S",
              ICB = "Natural Language Sentences (ICB Corpus)",
              ICBe = "Machine Readable Events (ICBe)"
             ) ) %>%

                      flextable::width( j = 1, width=0.25)  %>% 
                      #flextable::width( j = 2, width=0.25)  %>% 
                      flextable::width( j = 2, width=8.0)  %>% 
                      flextable::width( j = 3, width=8.0)  %>% 
                      flextable::fontsize(size = 9, part = "all") %>%
                      valign(j = 1:3, valign = "top", part = "body") %>%
                      flextable::line_spacing( space = 1.0, part = "all") %>% 
                      padding( padding = 1, part = "all") %>% 
                      bg( i = which( temp$banding==1 ) , part = "body", bg = "#EFEFEF", j=1:3) %>%
                      ftExtra::colformat_md(j = 3, part="body")
#ft_cuba_precision

ft_cuba_precision %>% save_as_image(path=here::here("replication_paper","pnas_draft","case_study_cuban_precision.png"), zoom=1) # webshot = "webshot2"
temp <- NULL

```

<!--- REX Remember to put in a Legend for the Icons ---->

\clearpage
\onecolumn
\begin{figure}[H]
\caption{Case Study 1: Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events \label{fig:case_study_cuban_precision}}
\centering{\includegraphics[ height=22cm]{"./case_study_cuban_precision.png"}}
\end{figure}
\clearpage
\twocolumn

Corpus  {.unnumbered}
========================================

For our corpus, $D$, we select a set of unusually high-quality historical narratives from the International Crisis Behavior (ICB) project ($n=471$)(SI Appendix, Table A1)[@brecherInternationalCrisisBehavior2017; @brecher_study_1997]. Their domain is 20th and 21st-century crises, defined as a change in the type, or an increase in the intensity, of disruptive interaction with a heightened probability of military hostilities that destabilizes states' relationships or challenges the structure of the international system [@brecherCrisesWorldPolitics1982].^[On near crises see @iakhnisCrisesWorldPolitics2019.] Crises are a significant focus of detailed single case studies and case comparisons because they provide an opportunity to examine behaviors in IR short of, or at least prior to, full conflict  [@holsti1914Case1965; @paigeKoreanDecisionJune1968; @allisonEssenceDecisionExplaining1971; @snyderConflictNationsBargaining1977; @gavinHistorySecurityStudies2014; @georgeDeterrenceAmericanForeign1974; @brecherCrisesWorldPolitics1982; @gaddisExpandingDataBase1987; @brecherPatternsCrisisManagement1988]. Case selection was exhaustive based on a survey of world news archives and region experts, cross-checked against other databases of war and conflict, and non-English sources [@kangUSBiasStudy2019; @brecherInternationalCrisisBehavior2017 p. 59]. Each narrative was written by consensus by a small number of scholars, using a uniform coding scheme, with similar specificity [@hewittEngagingInternationalData2001]. The corpus is unique in IR because it is designed to be used in a downstream quantitative coding project.

Prior Beliefs about IR, Ontological Coverage, and the Existing State of the Art {.unnumbered}
========================================

Next we draw informative prior beliefs about the underlying process of IR that we expect to govern behavior during historical episodes and their conversion to the historical record. We organize our prior beliefs along two overarching axes, summarized in detail by Table 1.

The first axis (rows) represents the types of information we expect to find in IR and forms the basis for our proposed ontology. We employ a metaphor of a chess game, with players (polities, rebel groups, IGOs, etc.), pieces (military platforms, civilians, domains), and behaviors (think, say, do). Precise sequencing is required to capture gambits (sequences of moves) and outcomes (victory, defeat, peace, etc.), while precise geo-coding is required to understand the chessboard (medium of conflict). We find `r codings_long_agreement %>% filter(varname_normalized %>% str_detect("actor")) %>% pull(value_normalized) %>% unique() %>% length()` actors and 117 different behaviors, and provide a full codebook in the online material.^[See the Github Repository [ICBEventData](https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$).]

We base our informed priors primarily on two sources of information. The first is the extensive existing measurement efforts of IR which we provide citations to alongside each concept. Second, we performed preliminary natural language processing of the corpus and identified named entities and behaviors mentioned in the text. Verbs were matched to the most likely definition found in Wordnet [@millerWordNetLexicalDatabase1995], tallied, and then aggregated into a smaller number hypernyms balancing conceptual detail and manageable sparsity for human coding (SI Appendix, Table A2).

The second axis (columns) compares the very high ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Additional relevant but dated or too small of an overlap in domain include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000], Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018], and COfEE [@balaliCOfEEComprehensiveOntology2021] (see histories in @merrittMeasuringEventsInternational1994 and @schrodtTwentyYearsKansas2006).] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010]. With the exception of large scale CAMEO dictionary based systems, the existing state of the art quantitative datasets ignore the vast majority of the information content found in international relations.^[See [@balaliCOfEEComprehensiveOntology2021] for a recent review of ontological depth and availability of Gold Standard example text.]

\clearpage
\onecolumn

```{r litreview, eval=T, echo=F, results='asis', include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

#target_file <- paste0(here::here(),"/replication_paper/data/in/icbe_litreview_trees_sentences.xlsx")
#lit_review_pull    <- readxl::read_excel(target_file, sheet="DatasetsComparisonLitReview")

library(googlesheets4) #Do this high up so you get the prompt early
library(lubridate, warn.conflicts = FALSE)
lit_review_pull <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/14RHRKgQGHj7Zr4_xt4fq2bIE7mDNayLKj89tYmOOfHQ/edit?usp=sharing", 
                                sheet="DatasetsComparisonLitReview") #%>% janitor::clean_names() 
dim(lit_review_pull)

library(flextable)
library(ftExtra)
library(tidyverse)
flextable::set_flextable_defaults(fonts_ignore=TRUE)

#https://davidgohel.github.io/flextable/reference/knit_print.flextable.html#chunk-options
# NOTE: If one of the citations is degenerate this will throw an error when knitting that is annoying to track down. You have to remove cite by cite to see which one it is.
set_flextable_defaults(fonts_ignore=TRUE)

lit_review_alt <- lit_review_pull %>% 
                  mutate(Literature =   paste0(" [",Literature %>% str_replace("; {0,}$","") ,"]")) %>%
                  mutate(Concept= Concept %>% paste0(Literature)) %>%
                  mutate(Concept = Concept %>% str_replace(fixed("[NA]"),"") ) %>%
                  dplyr::select(-Codebook, -`Cameo Codes`, -`ICB Codes`, -Literature, -`ICB Corpus`) 
                  #dplyr::filter(is.na(group))

n=nrow(lit_review_alt)
k=ncol(lit_review_alt)
horizontal = c(1,k)
r_intro= lit_review_pull$...1 %in% "Domain" %>% sum() #changed to domain
r_players= ( lit_review_pull$...1 %in% "Players" %>% sum() ) + r_intro
r_pieces= ( lit_review_pull$...1 %in% "Pieces" %>% sum() ) + r_players
r_think= ( lit_review_pull$...1 %in% "Think" %>% sum() ) + r_pieces
r_say= ( lit_review_pull$...1 %in% "Say" %>% sum() ) + r_think
r_unarmed= ( lit_review_pull$...2 %in% "Unarmed" %>% sum() ) + r_say
r_armed= ( lit_review_pull$...2 %in% "Armed" %>% sum() ) + r_unarmed
ft <- NULL

library(officer)
big_border = fp_border(color="black", width = 2)

bands=which( 1:nrow(lit_review_alt) %% 2 == 1)
litreview_ft_pnas <- lit_review_alt %>%
      flextable::as_flextable() %>%

      #add_header( 
      #        #"ICB Corpus"="Corpus",
      #        "ICBe"="Ours",
      #        "Phoenix"="CAMEO Dictionary\nSystems",
      #        "Terrier"="CAMEO Dictionary\nSystems",
      #        "ICEWs"="CAMEO Dictionary\nSystems",
      #        "MIDs Events"="Disputes",
      #        "UCDP-GED"="Violence",
      #        "ICB"="Crises",
      #        "COW"="Wars",
      #        "MIDs"="Disputes",
      #        top = T ) %>%

      add_header( 
                  #"ICB Corpus"="Corpus",
                  "ICBe"="Events Datasets",
                  "Phoenix"="Events Datasets",
                  "Terrier"="Events Datasets",
                  "ICEWs"="Events Datasets",
                  "MIDs Incidents"="Events Datasets",
                  "UCDP-GED"="Events Datasets",
                  "ICB"="Episodes Datasets",
                  "COW"="Episodes Datasets",
                  "MIDs"="Episodes Datasets",
                  top = T ) %>%

      set_header_labels(
        values = list("...1" = "" , "...2" = "" ) 
        ) %>%

      merge_h(part = "header") %>% #it's this merge that's killing us

      #Global Borders
      flextable::border_remove() %>% 
      #ok so because of our horizontal merging of headers, we have to manually place a right bar at j=10 2 short of the full width
      flextable::border_outer( part="all", border = big_border ) %>% #this will have a gap
      border(j=10:k, i=1 , part="header", border.right = big_border ) %>% #this fills the gap
      #There's another gap on the bottom left
      border(j=1, i=(n-15):n , part="body", border.bottom =  big_border ) %>% #this fills the gap
      border(j=2, i=(n-5):n , part="body", border.bottom =  big_border ) %>% #this fills the gap

      flextable::line_spacing( space = 1.0, part = "all") %>% #i=1:nrow(lit_review_alt),
      flextable::padding(padding = 0, part = "body") %>%
      flextable::fontsize(size = 7, part = "all") %>%
      #Banding
      bg( i = bands  , j=3:k, part = "body", bg = "#EFEFEF") %>%

      flextable::rotate(i=2, , j=3:k, rotation="tbrl",part="header") %>%
      flextable::align(align = "center", part = "body") %>%
      flextable::align(align = "center", part = "header")   %>%

      flextable::colformat_char(
          j = 1:ncol(lit_review_alt),
          na_str = "",
          prefix = "",
          suffix = ""
      ) %>%
      #Domains
      hline(i=c(r_intro), border = NULL, part = "body") %>% 
      merge_at(i = 1:r_intro, j = 1:2, part = "body") %>%  
      merge_at(i = 4, j = 5:7, part = "body") %>%
      merge_at(i = 5, j = 5:7, part = "body") %>%
      merge_at(i = 6, j = 5:7, part = "body") %>%
      merge_at(i = 7, j = 5:7, part = "body") %>%

      #players
      hline(i=c(r_players), border = NULL, part = "body") %>% 
      merge_at(i = (r_intro+1):r_players, j = 1:2, part = "body") %>%  
      #pieces
      hline(i=c(r_pieces), border = NULL, part = "body") %>% 
      merge_at(i = (r_players+1):r_pieces, j = 1:2, part = "body") %>%  
      #think
      hline(i=c(r_think), border = NULL, part = "body") %>% 
      merge_at(i = (r_pieces+1):r_think, j = 1:2, part = "body") %>%  
      #say
      hline(i=c(r_say), border = NULL, part = "body") %>% 
      merge_at(i = (r_think+1):r_say, j = 1:2, part = "body") %>%
      #Do
      merge_at(i = (r_say+1):r_armed, j = 1, part = "body") %>%

      #n_unarmed
      hline(i=c(r_unarmed), border = NULL, part = "body") %>% 
      merge_at(i = (r_say+1):r_unarmed, j = 2, part = "body") %>%  
      #n_armed
      #hline(i=c(r_armed), border = NULL, part = "body") %>% 
      merge_at(i = (r_unarmed+1):r_armed, j = 2, part = "body") %>%

      #set_table_properties(layout="autofit")  %>%
      vline( i = 1:2, j = c(3,9), border = NULL, part = "header") %>%
      vline( i = 1:(n), j = 1:2, border = NULL, part = "body")  %>%
      vline( i = 1:(n), j = c(3,9), border = NULL, part = "body")  %>%

      flextable::width( j = 3, width=3) %>% 
      flextable::width( j = (4):k, width=0.25) %>%
      flextable::line_spacing( space = 0.5, part = "body") %>% #i=1:nrow(lit_review_alt),
      flextable::padding(padding = 0, part = "body") %>%

      flextable::rotate(j=1, rotation="tbrl",part="all") %>%
      flextable::rotate(j=2, rotation="tbrl",part="all") %>%
      flextable::width( j = 1:2, width=0.1)  %>%
      ftExtra::colformat_md(j = 3, part="body")   #Make sure this goes last or it'll get overwritten by the colformat above

#https://cran.r-project.org/web/packages/ftExtra/vignettes/format_columns.html
#litreview_ft_pnas

#litreview_ft_pnas  %>% saveRDS(paste0(here::here(), '/replication_paper/tables/litreview_ft_pnas.Rds'))
litreview_ft_pnas %>% 
  ftExtra::colformat_md(j = 3, part="body") %>%
  flextable::set_caption(caption = "Ontological coverage of ICBe versus existing State of the Art") %>%
  # flextable::set_caption("Prior Beliefs, State of the Art, and Existing Data \\label{tbl:litreview}")  %>% 
  flextable::flextable_to_rmd()

```

\clearpage
\twocolumn

```{r , eval=T, results='hide', echo=F, include=T, message=F, cache=F}

n_crisis <- ICBe_events_agreed %>% dplyr::select(crisno) %>% unique() %>% nrow()
n_sentences <- ICBe_events_agreed %>% dplyr::select(crisno, sentence_number_int_aligned) %>% unique() %>% nrow()
#n_events <- ICBe_events_agreed %>% dplyr::filter(event_number_int %in% c(1,2,3)) %>% nrow()
#n_events_cuba <- ICBe_events_agreed %>% dplyr::filter(crisno == 196 & event_number_int %in% c(1,2,3)) %>% nrow()
n_coders <- icb_long_clean %>% dplyr::select(email_id) %>% unique() %>% nrow()

avg_coders_per_crisis <- icb_long_clean %>% dplyr::filter(crisno != 196) %>% dplyr::mutate(cris_coder = paste(crisno, email_id)) %>% dplyr::select(crisno, email_id) %>% unique() %>% dplyr::count(crisno) %>% dplyr::select(n) %>% lapply(mean)

avg_coders_per_sentence <- icb_long_clean %>% dplyr::filter(crisno != 196) %>% dplyr::mutate(cris_sent = paste(crisno, sentence_number_int_aligned)) %>% dplyr::select(cris_sent, email_id) %>% unique() %>% dplyr::count(cris_sent) %>% dplyr::select(n) %>% lapply(mean)

```

ICBe Coding Process {.unnumbered}
========================================

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].^[This process quickly focuses the coder on a smaller number of relevant options while also allowing them to apply multiple tags if the sentence explicitly includes more than one or there is insufficient evidence to choose only one tag. The guided coding process also allows for the possibility that earlier coarse decisions have less error than later fine-grained decisions.] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (SI Appendix, Fig. B1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players ($P$), whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more non-mutually exclusive specific activities. Some additional details like location and timing information was always collected while other details were only collected if appropriate, e.g. force size, fatalities, domains, units, etc. A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions.

Each crisis was typically assigned to 2 expert coders and 2 novice coders with an additional tie-breaking expert coder assigned to sentences with high disagreement.^[Expert coders were graduate students or postgraduates who collaboratively developed the ontology and documentation for the codebook. Undergraduate coders were students who engaged in classroom workshops.] For the purposes of measuring intercoder agreement and consensus, we temporarily disaggregate the unit of analysis to the Coder-Crisis-Sentence-Tag (n=`r icb_long_clean %>% nrow() %>% scales::comma()`), where a tag is any unique piece of information a coder can associate with a sentence such as an actor, date, behavior, etc. We then aggregate those tags into final events (n=`r ICBe_events_agreed %>% nrow() %>% scales::comma()`), using a consensus procedure (SI Appendix, Algorithm B2) that requires a tag to have been chosen by at least one expert coder and either a majority of expert or novice coders. This screens noisy tags that no expert considered possible but leverages novice knowledge to tie-break between equally plausible tags chosen by experts.

```{r}
#We calculate means in the SI appendix on agreement
means_group_any <- codings_long_agreement %>% filter(value_normalized!='') %>% 
                   group_by(crisno,sentence_number_int_aligned, varname_normalized) %>% arrange(desc(percent_expert)) %>% filter(row_number()==1) %>% ungroup() %>% 
                   dplyr::select(Experts=percent_expert,Novices=percent_novice) %>% summarise_all(mean)

means_group_keep <- codings_long_agreement %>% filter(value_normalized!='') %>% 
                    filter(algo1==1) %>%
                    group_by(crisno,sentence_number_int_aligned, varname_normalized) %>% arrange(desc(percent_expert)) %>% filter(row_number()==1) %>% ungroup() %>% 
                    dplyr::select(Experts_keep=percent_expert,Novices_keep=percent_novice) %>% summarise_all(mean)
```

Internal Consistency {.unnumbered}
========================================

We evaluate the internal validity of the coding process in several ways. For every tag applied we calculate the observed intercoder agreement as the percent of other coders who also applied that same tag (SI Appendix, Fig. B3). Across all concepts, the Top 1 Tag Agreement was low among novices (`r means_group_any$Novices %>% round(2) %>% scales::percent()`), moderate for experts (`r means_group_any$Experts %>% round(2) %>% scales::percent()`), and high (`r means_group_keep$Experts_keep %>% round(2) %>% scales::percent()`) following the consensus screening procedure.

We attribute the remaining disagreement primarily to three sources. First, we required coders to rate their confidence which was observed to be low for 20% of sentences- half due to a mismatch between the ontology and the text ("survey doesn't fit event"-45%) and half due to a lack of information or confused writing in the source text ("more knowledge needed"-40%, "confusing sentence"-6%). Observed disagreement varied predictably with self reported confidence (SI Appendix, Fig. B4). Second, as intended agreement is higher (75-80%) for questions with fewer options near the root of the ontology compared to agreement for questions near the leafs of the ontology (50%-60%). Third, individual coders exhibiting nontrivial coding styles, e.g. some more expressive applying many tags per concept while others focused on only the single best match. We further observed unintended synonymity, e.g. the same information can be framed as either a threat to do something or a promise not to do something.

Case Study Selection {.unnumbered}
========================================

The remaining two qualities we seek to measure are recall and precision of ICBe events in absolute terms and relative to other existing systems. We provide full ICB narratives, ICBe coding in an easy to read icongraphic form, and a wide range of visualizations for every case on the companion website. In this paper, we focus on two deep case studies. The first is the Cuban Missile Crisis (Figure 1) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war (hereafter Cuban Missiles). The second is the Crimea-Donbas Crisis (SI Appendix Figure D1) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full scale invasion (herafter Crimea-Donbas). Both cases involve a superpower in crisis with a neighbor, initiated by a change from a friendly to hostile regime, with implications for economic and military security for the superpower, risked full scale invasion, and eventually invited intervention by opposing superpowers. We choose these cases because they are substantively significant to 20th and 21st century international relations, widely known across scientific disciplines and popular culture, and are sufficiently brief to evaluate in depth.

Recall {.unnumbered}
========================================

Recall measures the share of desired information recovered by a sequence of coded events, $Pr(E|H)$, and is poorly defined for historical episodes. First, there is no genuine ground truth about what occurred, only surviving texts about it. Second, there is no \textit{a priori} guide to what information is necessary detail and what is ignorable trivia. History suffers from what is known as the Coastline Paradox [@mandelbrotFractalGeometryNature1983] --- it has a fractal dimension greater than one such that the more you zoom in the more detail you will find about individual events and in between every two discrete events. The ICBe ontology is a proposal about what information is important, but we need an independent benchmark to evaluate whether that proposal is a good one and that allows for comparing proposals from event projects that had different goals. We need a yardstick for history.

>>>>

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
#pnas_type: pnasresearcharticle

bibliography: ICBintro.bib
#csl: pnas.csl

@@

#https://bookdown.org/yihui/bookdown/a-single-document.html
#https://stackoverflow.com/questions/52531637/knitr-rmarkdown-latex-how-to-cross-reference-figures-and-tables-in-2-different/52532269#52532269
#https://stackoverflow.com/questions/51595939/bookdown-cross-reference-figure-in-another-file
#https://stackoverflow.com/questions/25824795/how-to-combine-two-rmarkdown-rmd-files-into-a-single-output/51521542#51521542
format:
  pdf: 
    latex_engine: xelatex
    citation_package: biblatex
    keep-tex: true
    latex-clean: false
    cite-method: biblatex

---
